Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-04-01 analysis/paper.tex
Add references

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. Modelled ranges do not
reliably predict the observed occurrence of specific species at
archaeological sites, but are aligned with more general trends in the
relative abundance of species.
\end{abstract}

@@

The Pleistocene--Holocene transition in West Asia marked a turning point
in global environmental history, as humans brought the first plants
under cultivation and began modifying surrounding ecosystems to support
their own subsistence. West Asia is part of the native range of a
remarkable number of domesticable plant species, including wild
relatives of wheat, barley, peas, lentils, and other crops of global
importance \citep{Diamond2002}. These species supported uniquely dense
and complex Late Epipalaeolithic (15--11.7 ka) societies based on
intensive foraging \citep{BarYosef1998, MaherEtAl2012} and eventually
plant management and pre-domestication cultivation \citep{Harris2007}.
As they were brought under domestication in the Pre-Pottery Neolithic
period (11.7--8.5 ka), the world's first agriculture was shaped by the
ecosystems from which it emerged and was embedded in.

Decades of research in archaeobotany and zooarchaeology have
reconstructed the subsistence economies of Late Epipalaeolithic and
Neolithic sites in great detail. Together with studies of charcoal,
pollen, soil, and a variety of palaeoclimate archives
\citep{JonesEtAl2019}, they also tell us much about the environments
surrounding these settlements. However, each of these sources of
evidence is subject to the wide variety of taphonomic and recovery
biases that are inherent in any direct record of the past. They are
also, by definition, records of the (human) environment at particular
times and places. Interpolating these snapshots to give a holistic
picture of the regional ecologies is not straightforward -- to date, it
has tended to rely on non-explicit, inductive modelling. The majority
are also filtered through human action, producing a mixed single that
makes it difficult to disentangle anthropic effects from the background
of environmental change in this period of rapid climatic alteration.

@@

\section{Background}\label{background}

\begin{itemize}
\tightlist
\item
  The transition to agriculture in West Asia was\ldots{}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Has always been important in study of agricultural origins

  \begin{itemize}
  \tightlist
  \item
    Historically: Vavilov, Pumpelly \& Childe
  \item
    Genetic studies tell us origin points, but not ranges
  \end{itemize}
\item
  Important to e.g.

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    \citep{MartinEtAl2016, MartinEtAl2025}
  \item
    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

  \begin{itemize}
  \tightlist
  \item
    More than the ``founder crops'';
  \item
    More than food
  \item
    (In archaebot., not all intentionally collected)
  \item
    Geographically and temporally diverse
  \item
    \ldots so we model lots of species!
  \end{itemize}
\item
  Regional ecological reconstructions generally rely on the `expert
  interpolation' (or what do they call it with isoscapes?) method

  \begin{itemize}
  \tightlist
  \item
    See CSEAS (AEA-prep) presentation
  \item
    Figure: comparisons
  \end{itemize}
\end{itemize}

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 97
taxa (Table~\ref{tbl-results-summary}) comprising the identifiable
species observed at at least two Late Epipalaeolithic/Pre-Pottery
Neolithic sites, according to our previous study of the regional
archaeobotanical data \citep{ArranzOtaeguiRoe2023}. This was based on
dataset collated from three previously published regional
archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
\citep{ORIGINS}, and COMPAG
\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
attempt to distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

@@

Our reconstructed palaeodistributions (shown in full in the appendix)
indicate that the majority of species have experienced a significant
reduction and/or shift in their ranges since the Early Holocene. 75 of
82 species have reduced ranges; 66 of more than 10\% or more. Though the
magnitude of the reduction from prehistory to the present likely
reflects a degree of overfitting in the model (discussed further in
Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled range
size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
(12.9--11.7 ka), and Early Holocene (11.7--8.3 ka) are more directly
comparable (Figure~\ref{fig-pred-range-density}). The average range of
modelled species was 16\% in the Early Holocene compared to the
Bølling--Allerød, and 25\% during the Younger Dryas (i.e.~ranges
recovered slightly between the Younger Dryas and Early Holocene). This
perhaps indicates that although this period is considered one of
climatatic amelioration globally \citep{JonesEtAl2019}, the colder
conditions of the Pleistocene may have supported more extensive
plant-based economies in West Asia specifically.

@@

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene (Figure~\ref{fig-palaeodist-sum}). These
include \emph{Ficus carica} (fig); \emph{Hordeum} spp. (wild barleys);
\emph{Lathyrus aphaca} and \emph{L. sativus} (both marginally edible
legumes); \emph{Triticum aestivum compactum} (in the N. Levant),
\emph{T. monococcum aegilopoides}, \emph{T. durum}, and \emph{Triticum
urartu} (but not the other wheat progenitor, \emph{T. turgidum dicoccum}
-- see Section~\ref{sec-discuss-crops}); \emph{Aegilops speltoides}, but
not \emph{Aegilops tauschii} (goatgrasses); and \emph{Vicia} spp.
(vetches), including \emph{Vicia faba} (broad beans). most of Anatolia,
Northern Mesopotamia, and the Zagros Mountains in particular disappear
from the predicted ranges of these species, leaving the Levant and to a
lesser extent the Aegean and Cyprus as refugia.

@@

Loss of the Northern Mesopotamia--Anatolia region from the predicted
ranges of crop progenitors is interesting in light of the `golden
triangle' hypothesis
\citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away
from this hypothesis and towards a more geographically diverse origin
\citep{Asouti2006, FullerEtAl2011, ArranzOtaeguiEtAl2016}, but it
remains the area with some of the earliest clear evidence of
domestication \citep{ZoharyEtAl2012, KabukcuEtAl2021, UlasEtAl2024}.
Comparative genetics also points to the Northern Mesopotamia region as
the centre of diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But
since these studies are based on modern genomes, if the wild range of
these plants has, as our modelling suggests, shifted since the Early
Holocene, the apparent centre of diversity may have shifted with them.
Our reconstructions are consistent with the late arrival of intensive
plant-based foraging economies in this region (cf.~the Natufian of the
Levant), and more broadly there need not be a link between the core
\emph{wild} range of a plant and the core area of its domestication. A
scenario in which cultivation emerged at the edges of the ranges of
valuable plant resources---as a means of extend their natural niche---is
also plausible.

The near-absence of the Zagros in any predicted ranges is also
surprising, given mounting evidence that domestication took place just
as early in the eastern \emph{Mashriq} as it did in the west
\citep{Zeder2024}. We consider that the most likely explanation for this
is that our flora does not include the species that were most important
to plant subsistence in the east. Archaeobotanical data on Neolithic
sites in the Zagros is limited (compared to the Levant in particular)
due to a hiatus in field research there from the 1980s to early 2010s
\citep{Zeder2024}. Recent research
\citep{WhitlamEtAl2018, GonzalezCarreteroEtAl2023} indicates that plant
subsistence in this region was based on a distinct set of species,
compared to the Levant and Anatolia.

@@

Exceptions to the dominant trend of range reduction include \emph{Cicer
reticulatum} (wild chickpea), which has a relatively stable range
centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
(wild emmer wheat), which is predicted to have two limited ranges
centered around the Black Sea Coast of Anatolia and the Palmyra basin.
In the latter case, neither of these areas are part of the predicted
modern distribution of wild emmer (centered around the Caucasus and
Northern Mesopotamia), but it would be consistent with archaeological
evidence for early cultivation at sites in the Upper Euphrates
\citep{Willcox2024}.

@@

Almost all the cereal and legume crop progenitors we modelled are
predicted to have only been found in the Levant during the terminal
Pleistocene and Early Holocene (see appendix). Part of this may be do
with the fact that both our initial flora and training occurrence
dataset have a strong bias towards the southern Levant, but strikingly
the modelled \emph{current} ranges of these plants do tend to include
Anatolia and the Zagros, so this cannot be the only factor. One notable
exception is the wild ancestor of chickpea (\emph{Cicer reticulatum}),
which is predicted to have a distribution centered on Northern
Mesopotamia but encompassing much of the the `hilly flanks' (except the
southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}). Another is
rye (\emph{Secale cereale}), which is inferred to be primarily Western
Anatolian (Figure~\ref{fig-palaeodist-chickpea-rye}). This is perhaps
relevant to rye's unusual domestication history, as a crop of West Asian
origin that was apparently not first cultivated until much later than
the `founder crops', in Europe \citep{SchreiberEtAl2021}.

@@

Flax (\emph{Linum bienne}) is predicted to have had a highly
concentrated distribution in Cyprus and along the Mediterranean coast of
the southern Levant. This is consistent with its low ubiquity in
archaeobotanical assemblages \citep{ArranzOtaeguiRoe2023}, despite
conventionally being considered a `founder crop', and presumably implies
that its domestication was similarly geographically constrained. It is
the only unambiguous crop progenitor with such a restricted range,
though pistachio (\emph{Pistacia atlantica}) and clubrush
(\emph{Bolboschoenus maritimus}) are similarly constrained to the
Mediterranean coast (and North Africa, in the case of pistachio).

@@

Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation
(Figure~\ref{fig-palaeodist-barely-levant}). It also sees a marked
decline in the archaeobotanical record from the Early PPNA/Early PPNB
(where it was amongst the most common taxa) to the Late PPNB and Late
Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
atlantica}) shows similar trends, but it is less certain that this
species was cultivated in the Neolithic. Conversely, the various wild
wheat species native to West Asia show almost no response to
Pleistocene/Holocene climate change, even within the Levant, and in the
archaeobotanical record wheat displays the opposite trend to barley and
pistachio -- becoming gradually more abundant through the course of the
Neolithic and dominant by its end. We hypothesise that climate-linked
range shifts were a factor driving these changes in the apparent
economic important of different crops: the decreasing availability of
barley and pistachio in the terminal Pleistocene may have prompted
cultivation as a strategy to retain access to them, though in the long
run it made them less attractive as staple crops compared to the more
resilient wheat. However, these dynamics are not seen in the majority of
crop progenitors.

@@

The failure of our hindcast models to predict the occurrence of species
in archaeobotanical assemblages has several possible explanations. Since
they do accurately predict the test dataset , a likely culprit is
overfitting of the models to the present environment. This implies that
the modelled palaeodistributions should be seen as conservative
estimates or a minimal range. Another obvious flaw in our methodology is
that the time slices used for palaeoclimatic reconstruction are very
broad---each covering around two millennia---and therefore potentially
unrepresentative of the environment around sites at the specific time at
which they were occupied. The variable quality of the archaeological
test dataset, especially in terms of chronology, is also a plausible
factor.

@@

It is consistent with more ``macro'' trends such as the reduced range of
Hordeum and Pistacia correlating with its reduced abundance in the
archaeobotanical record.

@@

``All models are wrong'' \citep{Box1976}, but the performance of our
models on independent test datasets give confidence in its predictions,
which are plausible and consistent with broad-scale patterns in the
archaeological record.

>>>>

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. The models performance in
predicting the occurrence of specific species at archaeological sites is
highly variable, but in aggregate the predictions are coherent and align
with broad-scale trends in the ubiquity and relative abundance of
species in the archaeological record.
\end{abstract}

@@

The Pleistocene--Holocene transition in West Asia marked a turning point
in global environmental history, as humans brought the first plants
under cultivation and began modifying surrounding ecosystems to support
their own subsistence. West Asia is part of the native range of a
remarkable number of domesticable plant species, including wild
relatives of wheat, barley, peas, lentils, and other crops of global
importance \citep{HarlanZohary1966, Diamond2002, ZoharyEtAl2012}. These
species supported uniquely dense and complex Late Epipalaeolithic
(15--11.7 ka) societies \citep{BarYosef1998, MaherEtAl2012} based on
foraging \citep{HarrisHillman1989, Colledge2001, WeissEtAl2004} and
eventually plant management and pre-domestication cultivation
\citep{Colledge2001, WeissEtAl2006, Harris2007, WillcoxEtAl2008}. The
first agro-ecosystems emerged as these plants were domesticated in the
Pre-Pottery Neolithic period (11.7--8.5 ka) and were shaped by the
broader ecosystem in which it was embedded.

Decades of research in archaeobotany and zooarchaeology have
reconstructed the subsistence economies of Late Epipalaeolithic and
Neolithic sites in great detail. Together with studies of other
environmental archaeological records and a variety of palaeoclimate
archives \citep{JonesEtAl2019}, they also tell us much about the
environments surrounding these settlements. However, each of these
sources of evidence is subject to the wide variety of taphonomic and
recovery biases that are inherent in any direct record of the past. They
are also, by definition, records of the (human) environment at
particular times and places. Interpolating these snapshots to give a
holistic picture of the regional ecologies is not straightforward -- to
date, it has tended to rely on non-explicit, inductive modelling. The
majority are also filtered through human action, producing a mixed
single that makes it difficult to disentangle anthropic effects from the
background of environmental change in this period of rapid climatic
alteration.

@@

\section{Background}\label{sec-bg}

The transition to agriculture represents one of the most fundamental
changes in human history. West Asia is one of the regions where this
process has been studied in the most detail: decades of research have
traced the gradual development of a Neolithic way of life and the
changes that occurred in the plant species and their geographical
distribution as a result. Although archaeobotanical assemblages can be
biased due to issues of preservation, sampling, recovery techniques, and
lab procedures \citep{Dennel1976, HastorfPopper1988}---and although they
include not just food remains but plant resources that were used for
other purposes or arrived at the site accidentally
\citep{HastorfPopper1988}---large-scale studies have still revealed
coherent patterns in the exploitation of plants over time
\citep{ColledgeEtAl2004, ArranzOtaeguiEtAl2016}.

The possibility of an abrupt, geographically-constrained process of
plant domestication was proposed in the 1990s
\citep{HillmanDavies1990, HillmanDavies1992, HeunEtAl1997, OzkanEtAl2002}
and developed as an explanatory model in the 2000s
\citep{LevYadunEtAl2000, GopherEtAl2001, AbboEtAl2005}. As part of this
model, some authors
\citep{LevYadunEtAl2000, GopherEtAl2001, AbboEtAl2010, AbboEtAl2012}
argued that eight plant species, collectively referred to as `founder
crops' or the `Neolithic crop package' \citep{ZoharyHopf1988} were
selected and domesticated once, without any phase of pre-domestication
cultivation \citep[p.~177]{AbboEtAl2011}. This process could have been
rapid under strong artificial selection
\citep{HillmanDavies1990, HillmanDavies1992} and may have occurred in a
single region or `core area' -- generally located in southeast Turkey
\citep{LadizinskyAdler1976, HeunEtAl1997, OzkanEtAl2002, OzkanEtAl2005, MoriEtAl2003, LuoEtAl2007}.
From this single point of origin, it was supposed that domesticated or
semi-domesticated plants radiated outwards to other regions
\citep{AbboEtAl2006, KilianEtAl2007, OzkanEtAl2011}.

The `short gestation' paradigm was challenged by others
\citep{Helbaek1969, Harris1989, Kislev1989, Colledge2001, WeissEtAl2004, WillcoxEtAl2008, FullerEtAl2018}.
Helbæk \citep[in][]{Kirkbride1966} argued that before the appearance of
domesticated plants, a phase of cultivation of wild seeds must have
taken place. The existence of a phase of cultivation of morphologically
wild cereals or `pre-domestication cultivation' was identified in the
archaeological record through the study of plant domestication traits
such as grain size, shattering v. non-shattering rachises. The
archaeobotanical evidence shows that during the Pre-Pottery Neolithic A
(PPNA) cereals exhibited sizes similar to those recorded in domestic
species, but their dispersal mechanism was still the same as the one
present in morphologically-wild species \citep[i.e.~shattering,
see][]{Kirkbride1966, Kislev1989, HillmanEtAl2001, Colledge2001, WillcoxEtAl2008}.
This evidence suggested that wild cereal stands could have been
cultivated for as much as a thousand years before non-shattering
domestic forms became prevalent in the archaeological record
\citep{TannoWillcox2006, TannoWillcox2012, ArranzOtaeguiEtAl2016}.
Additional archaeobotanical
\citep{Colledge2001, WillcoxEtAl2008, WillcoxEtAl2009, RiehlEtAl2013, ArranzOtaeguiEtAl2016, WeideEtAl2018, DoucheWillcox2018, WhitlamEtAl2018}
and genetic data
\citep{BadrEtAl2000, MolinaCanoEtAl2005, KilianEtAl2007, OzkanEtAl2011, IobBotigue2023}
in recent years has further challenged the short-gestation model to
explain the origins of plant domestication and agriculture in West Asia.

Similarly, the concept of a limited set of eight `founder crops'
\citep{ZoharyHopf1988} that were the first species cultivated,
domesticated and then spread as the basis of Neolithic agricultural
systems, is not supported by the latest evidence. Our previous analyses
of the composition of available archaeobotanical datasets shows that
these crops were of marginal importance during the Epipalaeolithic
period \citep{ArranzOtaeguiEtAl2018} and that Neolithic subsistence did
not rely either solely or primarily on the exploitation of these species
\citep{ArranzOtaeguiRoe2023}. Instead, multiple species of grasses,
legumes, fruits, nuts, and other plants were exploited over the Late
Pleistocene-Early Holocene transition in southwest Asia.

@@

The study of the natural distribution of the progenitors of domesticated
crops has been a central part of discussions on the origins of
agriculture and plant domestication from the beginning.
\citet{Humboldt1807} acknowledged the importance of the natural
distribution of wild species to explain the origin and domestication of
crops like spelt and rye. \citet{Candolle1886} integrated the study of
plant ecology and biogeography and influenced \citet{Darwin1859}, who
later reflected in detail about the geographical distribution of plants
and species diversity. At that time, there was intense debate about
whether there were single or multiple ``centres of creation'' of
species. Researchers aimed to evaluate whether plant and animal species
emerged in the same locations where they were currently distributed.

After Darwin, this early interest in crop origins evolved into more
specific discussions about the ``centres of plant domestication''.
\citet{Vavilov1926} was among the first to seek to determine the number
of regions in which plants had been independently domesticated
\citep{Harris1990}. His main method was `differential phytogeography':
he classified the variation within a crop and established the regions of
maximum diversity, to locate the geographic regions in which crops
originated. Using this method, Vavilov suggested that there were at
least eight centres of origin. His work was later criticised by
\citet{Harlan1971}, who argued that `centres of origin' and `centres of
diversity' had to be separated. For Harlan, a `centre' was as an ``area
in which things originate and out of which things are dispersed''
\citep[p.~468]{Harlan1971}, and he suggested that three main centres of
origin of domesticated crops existed. He further indicated that
Vavilov's approach to the question was simplistic and that more data
proxies had to be considered (e.g.~archaeology, history, geology), an
approach more in the tradition of \citet{Candolle1886}. Indeed, the
inclusion of archaeobotany and genetics in the last decades, together
with the study of wild relative distributions has been fundamental in
characterising the origins of agriculture \citep{FullerColledge2008}. As
a result of modern interdisciplinary studies, the number of recognised
centres of plant domestication has increased considerably, from the
three centres suggested by Harlan in 1971 to the six to eight centres
argued for in the 1990s \citep{Smith1995} and up to as much as 24
potential centres reported in 2009
\citetext{\citealp{PuruggananFuller2009}; \citealp[see
also][]{Fuller2010}}.

Biogeographic research into the centres of origin and/or domestication
of crops has also long informed broader understanding of the process of
agricultural origins. In \emph{Man Makes Himself}, \citet{Childe1936}
correctly located the centre of origin of European agriculture in the
`Fertile Crescent' of West Asia (unlike for example \citet{Pumpelly1908}
before him). This was not based on the region's prehistoric
archaeological record, which at that point was barely known. Instead he
was guided to the region by biogeographic work by Vavilov and
\citet{PeakeFleure1927}; only later was this prediction validated by
archaeological work on the Epipalaeolithic and Neolithic of Palestine
\citep{Boyd2018}. In subsequent decades, the search for more precise
origin zones of specific domestic plants relied on the assumption that
``the locus of domestication of a wild plant would presumably be within
its area of original distribution in the wild state''
\citetext{\citealp{Butzer1971}; \citealp[paraphrasing][]{Helbaek1959}}
-- and that this ``natural habitat'' has not changed significantly over
the last 12,000 years \citep{Butzer1971}.

Contemporary research on crop origins was pioneered by Harlan and
Zohary, who compared the current distribution of the wild progenitors of
domesticated plants in southwest Asia
\citep{HarlanZohary1966, Zohary1969, Zohary1973, ZoharyHopf1973} to the
rapidly-expanding archaeobotanical record
\citetext{\citealp{Harlan1971}; \citealp{Harlan1977}; \citealp[see
also][]{ZoharyHopf1988}; \citealp{HarlanZohary1966}}. They both were
interested in evaluating which were the wild ancestors of domesticated
crops and studying their natural distribution to understand their
domestication process
\citep{Zohary1969, Zohary1973, ZoharySpiegelRoy1975}. Indeed, the
natural distribution of the wild relatives of domestic plant species was
later used as a criterion to infer `pre-domestic cultivation' in the
archaeological record. For example, the presence of seeds of chickpea at
Jericho led \citet{Hopf1986} to interpret the remains as cultivars, as
the natural distribution of the wild form of chickpea was located
further to the north. The same rationale was applied to the einkorn
remains found at several Pre-Pottery Neolithic sites in the southern
Levant \citep{Hopf1969, Colledge2001}, as the wild progenitors of this
species was thought to be restricted to the northern Levantine area
\citep{HeunEtAl1997, ZoharyEtAl2012}. The same idea---presence of plants
outside their natural range---has been repeated in the literature more
recently by several other authors
\citep{TannoWillcox2006, WillcoxEtAl2008, HillmanEtAl2001}.

Despite the importance of biogeography in the development and validation
of hypotheses regarding the origins of agriculture, there have been few
studies of the wild range of specific crop progenitors or other relevant
plant species \citep[cf.~for domestic animals, e.g.][]{YeomansEtAl2017}.
Observations regarding translocation or range expansion must therefore
rely on a relatively rough and ahistoric notion of a species' `natural
distribution' -- that is, one based primarily on contemporary or
recent-historic occurrences. Yet we know there has been considerable
climatic and environmental change in West Asia since the terminal
Pleistocene \citep{JonesEtAl2019}, so it is very unlikely that these
ranges were in fact static. Reconstructions of broader environments have
attempted to trace their fluctuations through time, either for the
entire region
\citetext{\citealp[e.g.][]{VanZeistBottema1991}; \citealp[Hillman
in][]{MooreEtAl2000}} or parts of it \citep[e.g.][]{Cordova2007}, but
these are at the level of the vegetation zone rather than individual
species. They also invariably rely on what might be called `expert
interpolation' (where the author composes a map based on his or her own
knowledge of the relevant data) rather than an explicit modelling
process. This makes it difficult, if not impossible, for users of such
reconstructions to understand exactly how they were derived or what
could explain, for example, the significant discrepencies between the
predictions of different experts.

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Based on current understandings, we
assume that plant-based subsistence during this period was broad,
geographically- and temporally-varied, and reflects a gradual,
geographically decentred, and nonlinear tranistion to greater reliance
on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
point was a list of 97 taxa (Table~\ref{tbl-results-summary}) comprising
the identifiable species observed at at least two Late
Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous
study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. This was based on dataset collated from
three previously published regional archaeobotanical databases: ADEMNES
\citep{ADEMNES}, ORIGINS \citep{ORIGINS}, and COMPAG
\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
attempt to distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

@@

Our reconstructed palaeodistributions (shown in full in the appendix)
indicate that the majority of species have experienced a significant
reduction and/or shift in their ranges since the Early Holocene. 75 of
82 species have reduced ranges; 66 of more than 10\% or more. Though the
magnitude of the change in range size from prehistory to the present
likely reflects a degree of overfitting in the model (discussed further
in Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled
range size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
(12.9--11.7 ka), and Early Holocene (11.7--8.3 ka) are more directly
comparable (Figure~\ref{fig-pred-range-density}). The average range of
modelled species was 16\% in the Early Holocene compared to the
Bølling--Allerød, and 25\% during the Younger Dryas (i.e.~ranges
recovered slightly between the Younger Dryas and Early Holocene). This
perhaps indicates that although this period is considered one of
climatatic amelioration globally \citep{JonesEtAl2019}, the colder
conditions of the Pleistocene may have supported more extensive
plant-based economies in West Asia specifically.

@@

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene (Figure~\ref{fig-palaeodist-sum}). These
include \emph{Ficus carica} (fig); \emph{Hordeum} spp. (wild barleys);
\emph{Lathyrus aphaca} and \emph{L. sativus} (both marginally edible
legumes); \emph{Triticum aestivum compactum} (in the N. Levant),
\emph{T. monococcum aegilopoides}, \emph{T. durum}, and \emph{Triticum
urartu} (but not the other wheat progenitor, \emph{T. turgidum dicoccum}
-- see Section~\ref{sec-discuss-crops}); \emph{Aegilops speltoides}, but
not \emph{Aegilops tauschii} (goatgrasses); and \emph{Vicia} spp.
(vetches), including \emph{Vicia faba} (broad beans). Most of Anatolia,
Northern Mesopotamia, and the Zagros Mountains in particular disappear
from the predicted ranges of these species, leaving the Levant and to a
lesser extent the Aegean and Cyprus as refugia.

@@

Loss of the Northern Mesopotamia--Anatolia region from the predicted
ranges of crop progenitors is interesting in light of the `golden
triangle' hypothesis
\citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
puts this region at the centre of the development of agriculture and
plant domestication. Multiple lines of archaeological evidence have
emerged that point away from this hypothesis and towards a more
geographically diverse origin
\citep{Asouti2006, FullerEtAl2011, ArranzOtaeguiEtAl2016}, and our
reconstructions are also consistent with the late arrival of intensive
plant-based foraging economies in this region (cf.~the Natufian of the
Levant).

The near-absence of the Zagros in any predicted ranges is also
surprising, given mounting evidence that animal domestication took place
just as early in the eastern \emph{Mashriq} as it did in the west
\citep{Zeder2024}. We consider that the most likely explanation for this
is that our flora does not include the species that were most important
to plant subsistence in the east. Archaeobotanical data on Neolithic
sites in the Zagros is limited (compared to the Levant in particular)
due to a hiatus in field research there from the 1980s to early 2010s
\citep{Zeder2024}. Recent research
\citep{RiehlEtAl2013, WeideEtAl2017, WeideEtAl2018, WhitlamEtAl2018, GonzalezCarreteroEtAl2023}
indicates that plant subsistence in this region was based on a distinct
set of species, compared to the Levant and Anatolia.

@@

Exceptions to the dominant trend of range reduction include \emph{Cicer
reticulatum} (wild chickpea), which has a relatively stable range
centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
(wild emmer wheat), which is predicted to have two limited ranges
centered around the Black Sea Coast of Anatolia and the Palmyra basin.
In the latter case, neither of these areas are part of the predicted (by
our model) modern distribution of wild emmer, which is centered around
the Caucasus and Northern Mesopotamia. But it would be consistent with
archaeological evidence for early cultivation at sites in the Upper
Euphrates \citep{Willcox2024}.

@@

Almost all the cereal and legume crop progenitors we modelled are
predicted to have only been found in the Levant during the terminal
Pleistocene and Early Holocene (see appendix). Part of this may be do
with the fact that both our initial flora and training occurrence
dataset have a strong bias towards the southern Levant, but strikingly
the modelled \emph{current} ranges of these plants do tend to include
Anatolia and the Zagros, so this cannot be the only factor. One notable
exception is the wild ancestor of chickpea (\emph{Cicer reticulatum}),
which is predicted to have a distribution centered on Northern
Mesopotamia but encompassing much of the the `hilly flanks' (except the
southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}). Another is
rye (\emph{Secale cereale}), which is inferred to be primarily Western
Anatolian (Figure~\ref{fig-palaeodist-chickpea-rye}). This is perhaps
relevant to rye's unusual domestication history, as a crop of West Asian
origin that was intensively exploited
\citep{HillmanEtAl2001, DoucheWillcox2018} but apparently not first
cultivated until much later than the `founder crops', in Europe
\citep{SchreiberEtAl2021}.

@@

Flax (\emph{Linum bienne}) is predicted to have had a highly
concentrated distribution in Cyprus and along the Mediterranean coast of
the southern Levant. This is consistent with its low ubiquity in
archaeobotanical assemblages \citep{ArranzOtaeguiRoe2023}, despite
conventionally being considered a `founder crop', and presumably implies
that its domestication was similarly geographically constrained. It is
the only unambiguous crop progenitor with such a restricted range,
though pistachio (\emph{Pistacia atlantica}) and clubrush
(\emph{Bolboschoenus maritimus}) are similarly constrained to the
Mediterranean coast (and North Africa, in the case of pistachio). This
is despite the fact that they are well-attested in the archaeotanical
record from across West Asia.

@@

Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation
(Figure~\ref{fig-palaeodist-barely-levant}). It also sees a marked
decline in the archaeobotanical record from the Early PPNA/Early PPNB
(where it was amongst the most common taxa) to the Late PPNB and Late
Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
atlantica}) shows similar trends, but it is less certain that this
species was managed in the Neolithic. Conversely, the various wild wheat
species native to West Asia show almost no response to
Pleistocene/Holocene climate change, even within the Levant, and in the
archaeobotanical record wheat displays the opposite trend to barley and
pistachio -- becoming gradually more abundant through the course of the
Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}. We
hypothesise that climate-linked range shifts were a factor driving these
changes in the apparent economic important of different crops: the
decreasing availability of barley and pistachio in the terminal
Pleistocene may have prompted cultivation as a strategy to retain access
to them, though in the long run it made them less attractive as staple
crops compared to the more resilient wheat. However, these dynamics are
not seen in the majority of crop progenitors and may also have involved
other factors, such as changes in cultural food preferences.

@@

The failure of our hindcast models to predict the occurrence of species
in archaeobotanical assemblages has several possible explanations. Since
they do accurately predict the test dataset, a likely culprit is
overfitting of the models to the present environment. This implies that
the modelled palaeodistributions should be seen as conservative
estimates or a minimal range. Another obvious flaw in our methodology is
that the time slices used for palaeoclimatic reconstruction are very
broad---each covering around two millennia---and therefore potentially
unrepresentative of the environment around sites at the specific time at
which they were occupied. The variable quality of the archaeological
test dataset, especially in terms of chronology, is also a plausible
factor.

@@

@@

``All models are wrong'' \citep{Box1976}, but the performance of our
models on independent test datasets give confidence in its predictions,
which are plausible and consistent with broad-scale patterns in the
archaeological record. Results with regard to specific
archaeologically-attested species are not as promising, but this also
reflects the incomplete and coarsely temporally-resolved nature of the
archaeobotanical verification dataset. Discrepencies between the
modelled distribution, whether on the broad scale (e.g.~the restricted
geographic range of most species compared to their attestation in the
archaeological record), or relating to specific species (i.e.~false
positives and false negatives), suggest several avenues for future
investigation.
